The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with \textit{Gaia} Data and Beyond

### 2019-05-21 paper/ms.tex
Typographical changes

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant under certain assumptions and
are therefore used as quantitative labels of a star's orbit. We report a
subtle but important systematic error that is induced in the actions as a
consequence of local midplane variations expected for the Milky Way. This
error is difficult to model because it is non-Gaussian and bimodal, with
neither mode peaking on the null value. An offset in the vertical position of
the Galactic midplane of $\sim15\,\pc$ for a thin disk-like orbit or $\sim
120\,\pc$ for a thick disk-like orbit is sufficient to induce a $25\%$
systematic error in the vertical action $J_z$. In FIRE simulations of Milky
Way-mass galaxies, these variations are on the order of $\sim100\,\pc$ at the
solar circle. From observations of the mean vertical velocity variation of
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the Milky Way
midplane variations are of order $\sim60\textup{--}170\,\pc$\,--\,also
consistent with recent three-dimensional dust maps. Action calculations and
orbit integrations which assume the global and local midplanes are identical
are likely to include the induced errors we report, depending on the volume
considered. Variation in the local standard of rest or distance to the
Galactic center causes similar issues. The variation of the midplane must be
taken into account when performing dynamical analysis across the large regions
of the disk accessible to \textit{Gaia} and future missions.

@@

\section{Motivation} \label{sec:ref_frame}
We first demonstrate the significance to action computations of a systematic
offset in the determination of the Galactic midplane, distance to the Galactic
center, or LSR. We will find that such offsets are especially important for
disk-like orbits. The consequences we explore here may also arise from various
other systematic errors. For instance, the axisymmetric Galactic potential model
used in many works to compute actions may not be a good description of the true
potential --- or the parameters used may yield a potential that is
systematically incorrect outside an original fitted region. In this work, we
assume that the Galaxy is perfectly described by our model axisymmetric
potential, and simply explore the consequences of offsets in the Galactocentric
coordinate system.

@@

In this case, 
    the azimuthal action is
\citep[][Section~3.5.3b]{2008gady.book.....B}:
\beq\label{eq:Jphi_epi}
J_{\phi} = R_g v_c\text{,}
\eeq
and
    the radial and vertical actions are
\beq\label{eq:JR_Jz_epi}
\begin{split}
J_R &=  E_R/\kappa \\
J_z &=  E_z / \nu\text{,} 
\end{split}
\eeq
where $E_R$ and $E_z$ are the energy per unit mass in the radial and vertical
coordinates, respectively. Therefore,
\beq\label{eq:JR_Jz_epi_energy}
\begin{split}
J_R &= \frac{v_R^2 + \kappa^2 (R-R_g)^2}{2\kappa} \\
J_z &= \frac{v_z^2 + \nu^2 z^2}{2\nu}\text{.}
\end{split}
\eeq
Using Equations~\eqref{eq:orbits_epi} and \eqref{eq:orbits_vel_epi}, we can
simplify this:
\beq\label{eq:JR_Jz_epi_final}
\begin{split}
J_R &= \frac{\kappa A_R^2}{2} = \frac{v_{R,\text{max}}^2}{2\kappa} \\
J_z &= \frac{\nu A_z^2}{2} = \frac{v_{z,\text{max}}^2}{2\nu}\text{,}
\end{split}
\eeq
where the last equality in each line comes from the fact that
$v_{R,\text{max}} = \kappa A_R$ and $v_{z,\text{max}} = \nu A_z$.

@@

We consider the observer making a measurement, integrating an orbit, and
computing actions every 
    megayear 
using the prescription above. However, we
specify the star's starting position using the systematically offset
coordinate system.
Essentially we are shifting and then reintegrating at each point along the original orbit.
The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first 
  gigayear 
of the orbit in the upper panels of
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}---we perform
a $4\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95th minus 5th percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for
\thin{}, \thick{}, and \halo{} orbits (Table~\ref{tab:orbits}). See discussion in the
text and Figure~\ref{fig:Jz_hist} for the justification in using this to measure the magnitude of the
induced error. The left, center, and right panels show the result for $J_R$,
$J_{\phi}$, and $J_z$, respectively. The upper panels consider an offset in
$z$ and the lower panels consider an offset in $x$ (equivalently, an offset in
the solar radius). In some panels we also plot as dashed lines the epicyclic
prediction of the induced action error (Equation~\eqref{eq:Ji_err_mosttime}).
In the epicyclic approximation, a $z$ offset only induces an error in $J_z$
--- for all three orbits the epicyclic approximation is a good description of
the $J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The
error in $J_R$ is somewhat well-described for \thin{} and \thick{},
and a poor description for \halo{}. For $J_{\phi}$, the
epicyclic approximation is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

To allow for potential small inaccuracies in the determination of the original
fiducial coordinate system, we also subtract the best fit 
    curve of the form 
    \beq A \sin{\left(\phi + B\right)} + C \label{eq:fit-curve}\eeq 
from the midplane as a
function of azimuth to account for an overall tilt of the midplane (a
simplified version of the strategy described in
\citealt{2019ApJ...871..145A}). 
%mm [perhaps this would be better presented as a table] For
%$\text{A}$, the values are $-165$, $45$, and $8.8\,\pc$, for
%$\text{B}$ the values are $0.67$, $-0.088$, and $0.031\,\text{rad}$
%and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$ for
%\mi{}, \mf{}, and \mm{}, respectively. 
    For models (\mi{}, \mf{}, \mm{}) the best fit values are $A =
    (-165, 45, 8.8)\,\pc$, $B = (38, -5.0, 1.8)\,\text{deg}$,
    and $C = (-69, 19, -18)\,\pc$.
For the assumed solar radius of $8.2\,\kpc$, we can approximate the
angle offset 
    $\Delta \theta$ 
for the $z$-axis from the values of 
      $A$. We find for the same models $\Delta \theta = (1.15, 0.31,
      0.062)\,\text{deg}$.
These angle offsets are consistent with the
values given in \citet{2018arXiv180610564S} for the difference between the
$z$-axis as defined by the gas and stars.

@@

\subsection{Estimates of Milky Way Midplane Offsets 
      }

@@

Work
by the \textit{Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of 
oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}, which can be explained with models of Sagittarius-like encounters \citep{2013MNRAS.429..159G,2018MNRAS.481..286L,2019MNRAS.485.3134L}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\text{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.

@@

\subsubsection{Galactic Center Position}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.,][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}. A
recent measurement using near-infrared interferometry places $R_0$ at $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}, or a 
     0.4\% 
uncertainty.

>>>>

Orbital properties of stars, computed from their six-dimensional phase space
measurements and an assumed Galactic potential, are used to understand the
structure and evolution of the Galaxy. Stellar actions, computed from orbits,
have the attractive quality of being invariant under certain assumptions and
are therefore used as quantitative labels of a star's orbit. We report a
subtle but important systematic error that is induced in the actions as a
consequence of local midplane variations expected for the Milky Way. This
error is difficult to model because it is non-Gaussian and bimodal, with
neither mode peaking on the null value. An offset in the vertical position of
the Galactic midplane of $\sim15\,\pc$ for a thin disk-like orbit or $\sim
120\,\pc$ for a thick disk-like orbit is sufficient to induce a $25\%$
systematic error in the vertical action $J_z$. In FIRE simulations of Milky
Way-mass galaxies, these variations are on the order of $\sim100\,\pc$ at the
solar circle. From observations of the mean vertical velocity variation of
$\sim5\textup{--}10\,\kms$ with radius, we estimate that the Milky Way
midplane variations are of order $\sim60\textup{--}170\,\pc$\,---\,also
consistent with recent three-dimensional dust maps. Action calculations and
orbit integrations which assume the global and local midplanes are identical
are likely to include the induced errors we report, depending on the volume
considered. Variation in the local standard of rest or distance to the
Galactic center causes similar issues. The variation of the midplane must be
taken into account when performing dynamical analysis across the large regions
of the disk accessible to \textit{Gaia} and future missions.

@@

\section{Motivation} \label{sec:ref_frame}
We first demonstrate the significance to action computations of a systematic
offset in the determination of the Galactic midplane, distance to the Galactic
center, or LSR. We will find that such offsets are especially important for
disk-like orbits. The consequences we explore here may also arise from various
other systematic errors. For instance, the axisymmetric Galactic potential model
used in many works to compute actions may not be a good description of the true
potential\,---\,or the parameters used may yield a potential that is
systematically incorrect outside an original fitted region. In this work, we
assume that the Galaxy is perfectly described by our model axisymmetric
potential, and simply explore the consequences of offsets in the Galactocentric
coordinate system.

@@

In this case, 
the azimuthal action is
\citep[][Section~3.5.3b]{2008gady.book.....B}:
\beq\label{eq:Jphi_epi}
J_{\phi} = R_g v_c\text{,}
\eeq
and the radial and vertical actions are
\beq\label{eq:JR_Jz_epi}
\begin{split}
J_R &= E_R / \kappa \\
J_z &= E_z / \nu\text{,} 
\end{split}
\eeq
where $E_R$ and $E_z$ are the energy per unit mass in the radial and vertical
coordinates, respectively. Therefore,
\beq\label{eq:JR_Jz_epi_energy}
\begin{split}
J_R &= \frac{v_R^2 + \kappa^2 (R-R_g)^2}{2\kappa} \\
J_z &= \frac{v_z^2 + \nu^2 z^2}{2\nu}\text{.}
\end{split}
\eeq
Using Equations~\eqref{eq:orbits_epi} and \eqref{eq:orbits_vel_epi}, we can
simplify this:
\beq\label{eq:JR_Jz_epi_final}
\begin{split}
J_R &= \frac{\kappa A_R^2}{2} = \frac{v_{R,\text{max}}^2}{2\kappa} \\
J_z &= \frac{\nu A_z^2}{2} = \frac{v_{z,\text{max}}^2}{2\nu}\text{,}
\end{split}
\eeq
where the last equality in each line comes from the fact that
$v_{R,\text{max}} = \kappa A_R$ and $v_{z,\text{max}} = \nu A_z$.

@@

We consider the observer making a measurement, integrating an orbit, and
computing actions every 
megayear 
using the prescription above. However, we
specify the star's starting position using the systematically offset
coordinate system.
Essentially we are shifting and then reintegrating at each point along the original orbit.
The actions computed using the offset coordinate system
for each phase-space starting point are shown for the first gigayear 
of the orbit in the upper panels of
Figure~\ref{fig:one_orbit_wrong_ref}.\footnote{Occasionally the numerical
scheme fails and very large actions are reported by \texttt{gala}\,---\,we perform
a $4\sigma$ clip on each action to exclude such orbits, but this only excludes
a total of $5$ orbits out of the $1000$ considered for
Figure~\ref{fig:one_orbit_wrong_ref}. Some numerical artifacts remain, but the
vast majority of orbits are computed properly.}

@@

\begin{figure*}[htb!]
\begin{center}
\includegraphics[width=\textwidth]{fig/schmactions_many_orbits.pdf}
\end{center}
\caption{We report one half the 95th minus 5th percentile of the error
in the measured action ($\Delta J_i$) from coordinate system errors for
\thin{}, \thick{}, and \halo{} orbits (Table~\ref{tab:orbits}). See discussion in the
text and Figure~\ref{fig:Jz_hist} for the justification in using this to measure the magnitude of the
induced error. The left, center, and right panels show the result for $J_R$,
$J_{\phi}$, and $J_z$, respectively. The upper panels consider an offset in
$z$ and the lower panels consider an offset in $x$ (equivalently, an offset in
the solar radius). In some panels we also plot as dashed lines the epicyclic
prediction of the induced action error (Equation~\eqref{eq:Ji_err_mosttime}).
In the epicyclic approximation, a $z$ offset only induces an error in $J_z$\,---\,for all three orbits the epicyclic approximation is a good description of
the $J_z$ error. An $x$ offset induces an error in $J_R$ and $J_{\phi}$. The
error in $J_R$ is somewhat well-described for \thin{} and \thick{},
and a poor description for \halo{}. For $J_{\phi}$, the
epicyclic approximation is not a good description for any orbit.}
\label{fig:many_orbit_wrong_ref}
\end{figure*}

@@

To allow for potential small inaccuracies in the determination of the original
fiducial coordinate system, we also subtract the best fit curve of the form 
\beq \label{eq:fit-curve}
A \sin{\left(\phi + B\right)} + C 
\eeq 
from the midplane as a
function of azimuth to account for an overall tilt of the midplane (a
simplified version of the strategy described in
\citealt{2019ApJ...871..145A}). 
%mm [perhaps this would be better presented as a table] For
%$\text{A}$, the values are $-165$, $45$, and $8.8\,\pc$, for
%$\text{B}$ the values are $0.67$, $-0.088$, and $0.031\,\text{rad}$
%and for $\text{C}$ the values are $-69$, $19$, and $-18\,\pc$ for
%\mi{}, \mf{}, and \mm{}, respectively. 
For models (\mi{}, \mf{}, \mm{}) the best fit values are $A =
(-165, 45, 8.8)\,\pc$, $B = (38, -5.0, 1.8)\,\text{deg}$,
and $C = (-69, 19, -18)\,\pc$.
For the assumed solar radius of $8.2\,\kpc$, we can approximate the
angle offset 
$\Delta \theta$ 
for the $z$-axis from the values of 
$A$. We find for the same models $\Delta \theta = (1.15, 0.31, 0.062)\,\text{deg}$.
These angle offsets are consistent with the
values given in \citet{2018arXiv180610564S} for the difference between the
$z$-axis as defined by the gas and stars.

@@

\subsection{Estimates of Milky Way Midplane Offsets}

@@

Work
by the \textit{Gaia} collaboration confirmed these preliminary results on the
velocity and spatial scales of 
oscillation with clear spatial maps made
using DR2 data of median $v_z$ over a significant Galactic volume
\citep{2018A&A...616A..11G, 2019arXiv190209569F}, which can be explained with models of Sagittarius-like encounters \citep{2013MNRAS.429..159G,2018MNRAS.481..286L,2019MNRAS.485.3134L}. We also see in the FIRE
simulation that the vertical velocity variation as a function of azimuth is
$\sim5\textup{--}10\,\kms$ (Figure~\ref{fig:lsr_variations}), consistent with
these observations.

@@

\subsubsection{Galactic Center Position}
First, one must define the center of the Galaxy. This is usually taken to be
the location of the central supermassive black hole, Sagittarius~A\textsuperscript{*}
\citep[\sgra{}, e.g.,][]{2004ApJ...616..872R}. From stellar motions near
\sgra{}, the distance from the Sun to \sgra{}, $R_0$, can be
precisely measured \citep{2009ApJ...692.1075G, 2018AA...615L..15G}. A
recent measurement using near-infrared interferometry places $R_0$ at $8.178 \pm 0.035$~kpc
\citep{2019arXiv190405721A}, or a $0.4\%$ 
uncertainty.
